---
title: "SDS 192 Project 2"
subtitle: "SDS 192"
author: 
- name: Ayla Hauser
  affiliation: Smith College
- name: Tabitha Githinji
  affiliation: Smith College
date: "`r format(Sys.Date(), '%B %e, %Y')`"
editor: 
  mode: source
output: 
  html_document:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    theme: lumen
    df_print: paged
    code_folding: show
execute: 
  echo: False
  warning: False
---

```{r setup}
#| echo: False
#| warning: False

library(tidyverse)
library(fec16)
library(knitr)
```

```{r helper-chunk}
#| echo: False
#| warning: False

# Look at help file
#?contributions

# Note that the contributions data frame only contains the first 1000 rows:
#contributions

# Download full contributions data from web and overwrite contributions using:
# read_all_*() function.
# You will need to install the usethis package for this line of code to work
contributions <- read_all_contributions()

# Now contributions has full 887,823 rows
#contributions
```

# Contributions of Candidates in MA 
```{r analysis-1}
#| echo: False
#| warning: False

# Group member 1 work here!
# Selected by entity types of candidates and candidate committee, grouping and summing their transaction amounts, filtering and adding by their name, and binding all the separated totals back into a table. 
# The total amount of money each candidate and candidate committees are contributing in Massachusetts from 2015 to 2016. 

contributions_MA <- contributions |>
  filter(state == "MA" & city == "BOSTON") |>
  filter(entity_tp == "CAN" | entity_tp == "CCM") |>
  select(-c("amndt_ind", "rpt_tp", "zip_code", "other_id", "cand_id", "tran_id")) 

elizabeth_warren <- contributions_MA |>
  group_by(name) |>
  summarize(sum = sum(transaction_amt)) |>
  arrange(desc(sum)) |>
  filter(grepl("ELIZABETH", name)) |>
  summarize(elizabeth_warren = sum(sum))

group_contributions <- contributions_MA |>
  group_by(name)|>
  summarize(sum = sum(transaction_amt)) 

stephen_lynch <- group_contributions |>
  filter(grepl("STEPHEN", name)) |>
  summarize(stephen_lynch = sum(sum))

joe_kennedy <- group_contributions |>
  filter(grepl("KENNEDY", name)) |>
  summarize(joe_kennedy = sum(sum))

markey_committee <- group_contributions |>
  filter(grepl("MARKEY", name)) |>
  summarize(edward_markey = sum(sum))

```

```{r}
#| label: tbl-contributions
#| tbl-cap: "The total amount of money each candidate and candidate committees are contributing in Massachusetts from 2015 to 2016."


binded <-cbind(elizabeth_warren, joe_kennedy, markey_committee, stephen_lynch)

kable(binded, col.names = c("Elizabeth Warren", "Joe Kennedy", "Edward Markey", "Stephen Lynch"))
```


```{r analysis-2}
#| echo: False
#| warning: False


# Group member 2 work here!
#glimpse(campaigns)
campaigns <- campaigns
#glimpse(results_senate)
results_senate <- results_senate

```


```{r}
#| echo: False
#| warning: False

#join the senate results dataset with tghe campaign dataset
results_campaigns <- campaigns |>
  inner_join(results_senate, by = join_by(cand_id)) |>
  select(cand_id, cand_name, party, won, ttl_disb) |>
   group_by(won) |>
  filter(party == "DEM" | party == "REP") |>
   arrange(desc(ttl_disb), .by_group = TRUE)

```

# Total Disbursements in 2015-2016 Senate Races
```{r}
#| echo: False
#| warning: False
#| label: fig-contributions
#| fig-cap: "Amount of money that was disbursed by both winning and losing candidates in the 2015-2016 Senate races."


ggplot(data = results_campaigns,
       aes(x = won,
           y = ttl_disb,
           color = party)) +
  geom_jitter() +
  theme_minimal() +
  labs(title = "The Total Disbursements of Candidates and their Election Success",
       x = "Won(True or False)",
       y = "Total Disbursements ($)")
  
```


# Blog post

In the 2015-2016 election cycle, the Federal Election Commission (FEC) provided detailed data on campaign contributions from candidates, committees, and political organizations. For this analysis, we focused specifically on Boston, Massachusetts, where we aimed to explore the total amount of money spent by candidates and their committees during their campaigns. The primary question guiding this analysis was: How much did each candidate and their committees spend in Massachusetts, and did the amount of money spent influence their success in the elections?

Our first step was to filter the dataset to include only contributions from candidates and candidate committees in Massachusetts, excluding irrelevant entries such as unreported or unknown contributions. We then calculated the total transaction amounts for four major candidates: Joe Kennedy, Stephen Lynch, Elizabeth Warren, and Edward Markey. We found that Elizabeth Warren and Edward Markey, both U.S. Senators, had the highest contributions, with Warren receiving $44,500 and Markey receiving $68,000. In comparison, the two congressional candidates, Joe Kennedy and Stephen Lynch, had much lower amounts, with Kennedy receiving $40,000 and Lynch receiving $9,000. These results suggest that Senate candidates generally raised more money than their congressional counterparts, which led us to explore the broader relationship between campaign spending and electoral success.

To further investigate, we turned our attention to the relationship between campaign spending and election outcomes. We examined how much money was disbursed by both winning and losing candidates in the 2015-2016 Senate races. Our analysis revealed a clear pattern: candidates who spent more money were more likely to win their races. When analyzing the total disbursements, it was evident that candidates who lost tended to have lower spending, while those who won had substantial campaign expenditures. This correlation highlights the importance of funding in securing electoral victories, although it also raises questions about the fairness of a system that potentially favors wealthier candidates.

One key takeaway from this analysis is the significant role money plays in determining the success of political candidates. While campaign contributions are necessary to run competitive campaigns, the findings underscore potential ethical concerns when joining data from different sources. For instance, when merging datasets like contributions and election results, it’s crucial to ensure that data integrity is maintained, particularly when using information from different times or data frames. There’s also a risk of data being incomplete or biased, which could skew analysis and lead to misleading conclusions. Therefore, transparency and careful data handling are essential in drawing ethical and accurate insights from such complex datasets.

