---
title: "SDS 192 Mini-Project 1"
subtitle: "SDS 192"
author: 
- name: Tabitha Githinji
  affiliation: Smith College
- name: Ayla Hauser
  affiliation: Smith College
date: "`r format(Sys.Date(), '%B %e, %Y')`"
editor: 
  mode: source
output: 
  html_document:
    fig_width: 7
    fig_height: 6
    fig_caption: true
    theme: lumen
    df_print: paged
    code_folding: show
execute: 
  echo: False
  warning: False
---



```{r setup, include=FALSE}
library(tidyverse)
library(rscorecard)


sc_key("CzdI6sFVmg18jVgTvyPCm6NuXl2QXrG4s6OfQ70Q")

```

```{r import}
#| echo: False
#| warning: False

scorecard <- sc_init() |>
  sc_year(2018) |>                 #Note how we are looking at only 2021 data here!
  sc_filter(stabbr == "MA") |>     #Note how we are looking at only Massachusetts data here!
  #The line below shows variable selection (there are lots of variables)
  sc_select(unitid, instnm, city, highdeg, control, ugds, adm_rate, costt4_a, costt4_p, pcip27, pctfloan, admcon7, cdr3, wdraw_orig_yr2_rt) |>
  sc_get()
```

```{r recode}
#| echo: False
#| warning: False
#  Here's an example of how to recode the control variable

# We are renaming the column control_text from control
scorecard$control_text <-
  #The recode function does the work. It calls 1 "Public", 2 " Private nonprofit",etc. 
  recode(
    scorecard$control, 
    "1" = "Public", 
    "2" = "Private nonprofit", 
    "3" = "Private for-profit",
    .default = NA_character_
  )
```

# Distribution of Admission Rates Across College in MA(2018)
How do the cost of attendance, and admission rates differ between public and private colleges in Massachusetts for the year 2018?
```{r plot_1}
#| echo: False
#| warning: False
# Create plot 1 here
#This histogram displays the distribution of admission rates across colleges in Massachusetts for 2018. The x-axis represents the admission rates (in percentage), while the y-axis indicates the number of colleges within each admission rate range. Faceting by university type allows for a comparative view of how admission rates differ between public , private for profit and private nonprofit institutions, highlighting trends in accessibility."
ggplot(data = scorecard,
  aes(x = adm_rate)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "black") +
  labs(title = "Average Admission Rates of Colleges in MA (2018)",
       x = "Admission rate",
       y = "Number of Colleges",
       subtitle = " College Scorecard Institution-Level Data ",
       caption = "Mapped the admission rate",
       ) +
  
   facet_wrap(~ control_text)+
  theme_minimal()  
  
```

# Cost of Attendance for each Type of University/College

```{r plot_2}
#| echo: False
#| warning: False
#Mutation creating a cost of attendance that combines both costs in the data set and filters for when the cost is greater than zero.
#Jitter plot with the x axis equal to the categorical variable of the type of university and the y axis equal to the numerical variable of cost of attendance of universities. This was done to compare the cost of attendance for type of university. The alpha argument was used to show the concentration of points at a certain place of the graph. 

scorecard <- scorecard |>
  mutate(cost_attendance = (costt4_a + costt4_p)) |>
  filter(cost_attendance > 0)

ggplot(data = scorecard,
       aes(x = control_text,
       y = cost_attendance)) +
  geom_jitter(alpha = 0.6,
              ) +
  theme_minimal() +
  labs(x = "Type of University/College",
       y = "Cost of Attendance($)",
       caption = "Graphed Cost of Attendance for each Type of University/College",
       title = "Cost of Attendence for Schools in Massachussetts") 
```

# Blog post

In this blog post, we delve into the admission rates and costs of attendance for colleges in Massachusetts, using data from the College Scorecard, a comprehensive dataset produced by the U.S. Department of Education. This dataset provides a wealth of information on various metrics related to higher education institutions, including tuition costs, graduation rates, and admission statistics. For our analysis, we focus specifically on the year 2018, examining how admission rates and costs of attendance differ between public universities, private nonprofit universities, and private profit universities. Our findings are illustrated through two visualizations: a histogram depicting admission rates and a jitter plot showcasing costs of attendance.

The first visualization, a histogram, reveals important insights into the distribution of admission rates among Massachusetts colleges. From the histogram, we observe that public universities generally exhibit higher admission rates, clustering between 70% and 90%, while private nonprofit universities display a wider range of admission rates, with many institutions falling below 60%. This indicates that public universities tend to be more accessible to a broader demographic of applicants compared to their private counterparts. The faceting of the histogram by university type allows us to easily compare these patterns, highlighting the distinct roles that public and private institutions play in the higher education landscape.

In addition to admission rates, our second visualization—a jitter plot—provides a clearer picture of the costs of attendance for these colleges. The jitter plot illustrates that public universities typically have lower costs compared to private nonprofit and private profit institutions. By representing the cost of attendance on the y-axis and the type of university on the x-axis, the plot reveals that many public colleges offer a more affordable option for students. The use of jittering helps to visualize the concentration of costs within each category, showcasing the financial challenges that may be faced by students attending private colleges.

The key takeaway from our analysis is the stark difference in both admission rates and costs of attendance between public and private colleges in Massachusetts. Public universities generally provide higher admission rates and lower costs, making them more accessible to a wider range of students. In contrast, private nonprofit and private profit institutions often have more selective admission processes and higher tuition fees, which may limit access for some students. However, it is crucial to approach this data with ethical considerations in mind. One primary concern is the representation of marginalized student groups in the dataset. While the College Scorecard provides valuable insights, it may not fully capture the unique barriers faced by students from low-income backgrounds or underrepresented communities. Furthermore, the dataset relies heavily on quantitative metrics, which can overlook qualitative aspects of students’ experiences and the broader context of their educational journeys.

In summary, our visualizations of admission rates and costs of attendance among Massachusetts colleges underscore important trends in accessibility and affordability. As we analyze such datasets, it is essential to remain aware of the ethical implications and potential biases inherent in the data collection process, ensuring that our findings contribute to a more equitable understanding of higher education access and opportunities for all students.


